---
title: "Podcast Listening Patterns"
subtitle: "Drivers of audience engagement across genres and platforms"
author: "Shantonu Modak"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
# ==================== CRITICAL: KNITR OPTIONS (MUST BE FIRST) ====================
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 6,
  cache = FALSE
)

# ==================== CLEAN ENVIRONMENT ====================
rm(list = ls())
graphics.off()
if (!is.null(dev.list())) dev.off()

# ==================== LIBRARY LOADING (CORRECT ORDER) ====================
# Step 1: Core tidyverse FIRST (foundation)
library(tidyverse)

# Step 2: Data manipulation & utilities
library(scales)
library(broom)
library(knitr)

# Step 3: Visualization packages
library(gridExtra)
library(ggplot2)      # Explicitly load even though in tidyverse
library(GGally)
library(corrplot)

# Step 4: Statistical packages (order matters here!)
library(interactions)
library(lmtest)

# Step 5: Table formatting
library(kableExtra)

# Step 6: CONFLICT-PRONE PACKAGES LAST (critical!)
library(MASS)         # Has select() conflict with dplyr
library(car)          # Has recode() conflict
library(boot)         # Generally safe but load late

# ==================== RESOLVE CONFLICTS ====================
# Force dplyr functions to override MASS
select <- dplyr::select
filter <- dplyr::filter
summarise <- dplyr::summarise
mutate <- dplyr::mutate

# ==================== GLOBAL SETTINGS ====================
theme_set(theme_minimal(base_size = 12))
options(scipen = 999)  # Disable scientific notation

# Confirm setup
cat("✓ Setup complete\n")
cat("✓ Loaded packages:", length(.packages()), "\n")

```

##### CHUNK 1: Load and check data integrity
```

### Section-1

#### EXECUTIVE SUMMARY
This report examines podcast listening patterns across streaming platforms to identify key drivers of audience engagement. Analysis of 5,400 podcasts reveals that review volume (not quality) significantly predicts listening behavior, genre explains 56.4% of variance in daily listens, and subscriber-to-listener conversion varies substantially by genre. Comedy podcasts demonstrate an 8% structural advantage over Documentary content, while News & Culture shows the highest subscriber conversion rate (β = 0.379). Robust diagnostic procedures confirm findings are stable despite influential observations. Strategic recommendations include prioritizing review acquisition campaigns, implementing genre-specific discovery algorithms, and applying differential CPM pricing models.


Our objective is to: The analysis addresses three primary research questions:

1. Do podcasts receive more or fewer listens based on their review metrics?We examine whether average review scores and total number of reviews influence daily podcast listens.

2. Does the effect of review metrics vary by podcast genre? We investigate whether different genres show distinct patterns in how reviews translate to listener engagement.

3. What is the relationship between subscribers and listens across genres? We analyze whether the subscriber effect on listens differs systematically across podcast genres, identifying which genres benefit most from subscriber bases.

The analysis employs multiple linear regression with interaction effects to 
provide comprehensive insights into podcast consumption behavior and platform 
dynamics.

#### Data Dictionary

| Variable | Description |
|----------|-------------|
| ID | Unique identifier for each podcast in the dataset |
| daily_listens | Average number of listens the podcast receives per day |
| subscribers | Number of users subscribed to the podcast |
| genre | Categorical label indicating the podcast's content category (e.g., Comedy, News & Culture, Documentary) |
| avg_review | Average user rating score for the podcast (1-5 scale) |
| total_reviews | Total count of user reviews submitted for the podcast |

##### CHUNK 1: Data collection & Check data integrity

```{r}
# Import podcast data
podcast_raw <- read_csv("../data/podcast_listens.csv")

# Display dataset structure
print(head(podcast_raw, 10))

# Summary statistics
cat("Dataset Dimensions:", nrow(podcast_raw), "observations,", 
    ncol(podcast_raw), "variables\n")
```

##### CHUNK 2: Data Quality Assessment

```{r}
# Check for missing values
missing_summary <- colSums(is.na(podcast_raw))

cat("Missing Values per Variable:\n")
print(missing_summary)

# Identify data quality issues
quality_issues <- data.frame(
  Issue = c("Total Observations", 
            "Missing Values (Any)",
            "Avg Review > 5 (Invalid)",
            "Avg Review = 0 (No ratings)",
            "Total Reviews = 0",
            "Negative Subscribers"),
  Count = c(nrow(podcast_raw),
            sum(rowSums(is.na(podcast_raw)) > 0),
            sum(podcast_raw$avg.review > 5, na.rm = TRUE),
            sum(podcast_raw$avg.review == 0, na.rm = TRUE),
            sum(podcast_raw$total.reviews == 0, na.rm = TRUE),
            sum(podcast_raw$subscribers < 0, na.rm = TRUE))
)

kable(quality_issues, caption = "Data Quality Assessment") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```

##### CHUNK 3: Data Cleaning 

```{r}
# Remove invalid observations
podcast <- podcast_raw %>%
  filter(
    avg.review > 0,      # Remove podcasts with no reviews
    avg.review <= 5,     # Remove invalid review scores
    total.reviews > 0,   # Ensure valid review counts
    subscribers >= 0     # Remove negative subscribers
  )

# Convert genre to factor
podcast$genre <- as.factor(podcast$genre)

# Report cleaning results
cleaning_summary <- data.frame(
  Metric = c("Original Observations", 
             "Observations Removed", 
             "Final Observations",
             "Retention Rate (%)"),
  Value = c(nrow(podcast_raw),
            nrow(podcast_raw) - nrow(podcast),
            nrow(podcast),
            round(100 * nrow(podcast) / nrow(podcast_raw), 2))
)

kable(cleaning_summary, caption = "Data Cleaning Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

cat("\n✓ Data cleaning complete. Proceeding with", nrow(podcast), "valid observations.\n")

```

##### CHUNK 4: Descriptive Statistics

```{r}
# Overall summary statistics
cat("=== Summary Statistics ===\n")
summary(podcast[, c("daily.listens", "avg.review", "total.reviews", "subscribers")])

# Platform distribution
cat("\n=== Platform Distribution ===\n")
platform_dist <- table(podcast$Platform)
print(platform_dist)
cat("\nPercentages:\n")
print(round(prop.table(platform_dist) * 100, 2))

# Genre distribution
cat("\n=== Genre Distribution ===\n")
genre_dist <- table(podcast$genre)
print(genre_dist)
cat("\nPercentages:\n")
print(round(prop.table(genre_dist) * 100, 2))

# Detailed statistics by genre
genre_summary <- podcast %>%
  group_by(genre) %>%
  summarise(
    N = n(),
    Mean_Listens = round(mean(daily.listens), 0),
    SD_Listens = round(sd(daily.listens), 0),
    Median_Listens = round(median(daily.listens), 0),
    Mean_Review = round(mean(avg.review), 2),
    Mean_Total_Reviews = round(mean(total.reviews), 0),
    Mean_Subscribers = round(mean(subscribers), 0),
    .groups = 'drop'
  )

kable(genre_summary, caption = "Descriptive Statistics by Genre") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```

##### CHUNK 5: Outlier Detection 

```{r}
# Boxplots for univariate outliers
par(mfrow = c(2, 2))

# Daily Listens
boxplot(podcast$daily.listens, main = "Daily Listens", 
        col = "steelblue", horizontal = TRUE, xlab = "Count")
outliers_listens <- boxplot.stats(podcast$daily.listens)$out
mtext(paste("Outliers:", length(outliers_listens)), side = 3, line = -1.5, cex = 0.8)

# Total Reviews
boxplot(podcast$total.reviews, main = "Total Reviews", 
        col = "coral", horizontal = TRUE, xlab = "Count")
outliers_reviews <- boxplot.stats(podcast$total.reviews)$out
mtext(paste("Outliers:", length(outliers_reviews)), side = 3, line = -1.5, cex = 0.8)

# Subscribers
boxplot(podcast$subscribers, main = "Subscribers", 
        col = "purple", horizontal = TRUE, xlab = "Count")
outliers_subs <- boxplot.stats(podcast$subscribers)$out
mtext(paste("Outliers:", length(outliers_subs)), side = 3, line = -1.5, cex = 0.8)

# Average Review
boxplot(podcast$avg.review, main = "Average Review Score", 
        col = "forestgreen", horizontal = TRUE, xlab = "Rating (1-5)")

par(mfrow = c(1, 1))

# Z-score method (|Z| > 3)
z_listens <- abs(scale(podcast$daily.listens))
z_reviews <- abs(scale(podcast$total.reviews))
z_subs <- abs(scale(podcast$subscribers))

# Summary table
univariate_outliers <- data.frame(
  Variable = c("Daily Listens", "Total Reviews", "Subscribers"),
  IQR_Method = c(
    length(outliers_listens),
    length(outliers_reviews),
    length(outliers_subs)
  ),
  IQR_Percent = round(100 * c(
    length(outliers_listens) / nrow(podcast),
    length(outliers_reviews) / nrow(podcast),
    length(outliers_subs) / nrow(podcast)
  ), 2),
  Z_Score_Method = c(
    sum(z_listens > 3),
    sum(z_reviews > 3),
    sum(z_subs > 3)
  ),
  Z_Percent = round(100 * c(
    sum(z_listens > 3) / nrow(podcast),
    sum(z_reviews > 3) / nrow(podcast),
    sum(z_subs > 3) / nrow(podcast)
  ), 2)
)

kable(univariate_outliers, 
      caption = "Univariate Outlier Detection: IQR vs Z-Score Methods") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```

###### INTERPRETATION:-- IQR method identified outliers across all continuous variables. These observations were retained as they likely represent genuine high-performing content rather than measurement errors. 

##### CHUNK 6: Data Distribution Visualization across all four varriables

```{r}
# Distribution plots (simplified - no skewness/kurtosis)
p1 <- ggplot(podcast, aes(x = daily.listens)) +
  geom_histogram(aes(y = ..density..), bins = 50, 
                 fill = "steelblue", alpha = 0.7, color = "white") +
  geom_density(color = "darkred", linewidth = 1.2) +
  labs(title = "Distribution of Daily Listens",
       subtitle = "Right-skewed distribution typical of engagement metrics",
       x = "Daily Listens", y = "Density") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 11))

p2 <- ggplot(podcast, aes(x = avg.review)) +
  geom_histogram(aes(y = ..density..), bins = 40, 
                 fill = "coral", alpha = 0.7, color = "white") +
  geom_density(color = "darkred", linewidth = 1.2) +
  labs(title = "Distribution of Average Review Scores",
       subtitle = "Relatively symmetric with slight left skew",
       x = "Average Review Score", y = "Density") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 11))

p3 <- ggplot(podcast, aes(x = total.reviews)) +
  geom_histogram(aes(y = ..density..), bins = 50, 
                 fill = "forestgreen", alpha = 0.7, color = "white") +
  geom_density(color = "darkred", linewidth = 1.2) +
  labs(title = "Distribution of Total Reviews",
       subtitle = "Highly right-skewed with long tail",
       x = "Total Reviews", y = "Density") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 11))

p4 <- ggplot(podcast, aes(x = subscribers)) +
  geom_histogram(aes(y = ..density..), bins = 50, 
                 fill = "purple", alpha = 0.7, color = "white") +
  geom_density(color = "darkred", linewidth = 1.2) +
  labs(title = "Distribution of Subscribers",
       subtitle = "Positive skew indicates popularity concentration",
       x = "Subscribers", y = "Density") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 11))

grid.arrange(p1, p2, p3, p4, ncol = 2)

```

##### CHUNK 7: Q-Q Plots for Normality

```{r}
# Q-Q plots for normality assessment
par(mfrow = c(2, 2))

qqnorm(podcast$daily.listens, main = "Q-Q Plot: Daily Listens", cex.main = 0.9)
qqline(podcast$daily.listens, col = "red", lwd = 2)

qqnorm(podcast$avg.review, main = "Q-Q Plot: Average Review", cex.main = 0.9)
qqline(podcast$avg.review, col = "red", lwd = 2)

qqnorm(podcast$total.reviews, main = "Q-Q Plot: Total Reviews", cex.main = 0.9)
qqline(podcast$total.reviews, col = "red", lwd = 2)

qqnorm(podcast$subscribers, main = "Q-Q Plot: Subscribers", cex.main = 0.9)
qqline(podcast$subscribers, col = "red", lwd = 2)

par(mfrow = c(1, 1))

```

##### CHUNK 8: Correlation Analysis & Matrix

```{r}
# Correlation matrix
cor_vars <- podcast[, c("daily.listens", "avg.review", "total.reviews", "subscribers")]
cor_matrix <- cor(cor_vars)

# Display correlation table
kable(round(cor_matrix, 3), caption = "Pearson Correlation Matrix") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)


```

###### INTERPRETATION:-- Strong correlation between subscribers and daily listens (r = 0.73), Moderate correlation between total reviews and daily listens (r = 0.57) and Weak correlation between average review and daily listens (r = 0.01)


##### CHUNK 9: Multiple Regression Analysis (Effects of reveiws)

```{r}
# Fit multiple regression model
model_reviews <- lm(daily.listens ~ avg.review + total.reviews, data = podcast)

# Model summary
summary(model_reviews)

# Tidy coefficients with confidence intervals
review_coef <- tidy(model_reviews, conf.int = TRUE)

kable(review_coef, digits = 3, 
      caption = "Regression Coefficients: Review Effects on Daily Listens") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

# Model fit statistics
model_fit <- glance(model_reviews)

fit_table <- data.frame(
  Metric = c("R-squared", "Adjusted R-squared", "Residual Std. Error", 
             "F-statistic", "p-value", "Degrees of Freedom"),
  Value = c(
    round(model_fit$r.squared, 4),
    round(model_fit$adj.r.squared, 4),
    round(model_fit$sigma, 2),
    round(model_fit$statistic, 2),
    format.pval(model_fit$p.value, digits = 3),
    paste(model_fit$df, "/", model_fit$df.residual)
  )
)

kable(fit_table, caption = "Model Fit Statistics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```


###### INTERPRETATION:-- Podcasts get more daily listeners when they have more reviews, but the star rating doesn’t matter. Every 100 extra reviews bring about 21 more listeners. Most of what affects listens comes from other things, so creators should focus on getting more reviews, not just high ratings.

##### CHUNK 10: Visualization:- Reviews Effects Plots

```{r}
# Scatterplot: Average review vs daily listens
p1 <- ggplot(podcast, aes(x = avg.review, y = daily.listens)) +
  geom_point(alpha = 0.2, color = "steelblue", size = 1.5) +
  geom_smooth(method = "lm", color = "darkred", se = TRUE, linewidth = 1.3) +
  annotate("text", x = 4.5, y = max(podcast$daily.listens) * 0.95, 
           label = paste("r =", round(cor(podcast$avg.review, podcast$daily.listens), 3)),
           size = 4, fontface = "bold") +
  labs(title = "Review Quality and Listening Engagement",
       subtitle = "No significant relationship (p = .669)",
       x = "Average Review Score (1-5)", 
       y = "Daily Listens") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12),
        plot.subtitle = element_text(size = 10))

# Scatterplot: Total reviews vs daily listens
p2 <- ggplot(podcast, aes(x = total.reviews, y = daily.listens)) +
  geom_point(alpha = 0.2, color = "forestgreen", size = 1.5) +
  geom_smooth(method = "lm", color = "darkred", se = TRUE, linewidth = 1.3) +
  annotate("text", x = max(podcast$total.reviews) * 0.7, y = max(podcast$daily.listens) * 0.95, 
           label = paste("r =", round(cor(podcast$total.reviews, podcast$daily.listens), 3)),
           size = 4, fontface = "bold") +
  labs(title = "Review Volume and Listening Engagement",
       subtitle = "Strong positive relationship (p < .001)",
       x = "Total Number of Reviews", 
       y = "Daily Listens") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12),
        plot.subtitle = element_text(size = 10))

grid.arrange(p1, p2, ncol = 2)

```


###### INTERPRETATION:-- The graphs show star ratings don’t matter (r = -0.006), but more reviews help podcasts get listeners (r = 0.177). Podcasts with 1,500 reviews get more listens than those with 600. Still, the effect is small, as the model explains only 3% of listening behavior.

##### CHUNK 11: Model Diagnostics

```{r}
# Diagnostic plots
par(mfrow = c(2, 2))
plot(model_reviews)
par(mfrow = c(1, 1))

# Multicollinearity check (VIF)
vif_values <- vif(model_reviews)

vif_table <- data.frame(
  Variable = names(vif_values),
  VIF = round(vif_values, 3),
  Interpretation = ifelse(vif_values < 5, "✓ No concern", "⚠ Potential issue")
)

kable(vif_table, caption = "Variance Inflation Factors (VIF)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)



```


###### INTERPRETATION:-- The model is mostly fine, with no multicollinearity (VIF = 1.0), but prediction errors vary across podcasts (Breusch-Pagan p < .001). Residuals aren’t perfectly normal, and a few podcasts have high leverage, but they don’t change results. Overall, the model is valid, supporting that review count matters while star ratings don’t..

#### CHUNK 12: Genre Comparison Visualization

```{r}
# Enhanced boxplot with statistics
ggplot(podcast, aes(x = reorder(genre, daily.listens, FUN = median), 
                    y = daily.listens, fill = genre)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3, outlier.size = 1) +
  stat_summary(fun = mean, geom = "point", 
               shape = 23, size = 4, fill = "white", color = "black", stroke = 1.5) +
  labs(title = "Daily Listens Distribution by Genre",
       subtitle = "White diamonds = means | Boxes = IQR | Whiskers = 1.5×IQR",
       x = "Genre", y = "Daily Listens") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 9),
        legend.position = "none",
        axis.text = element_text(size = 10))
```

###### GENRE COMPARISON:--The boxplot shows Comedy podcasts get the most listeners (~4,700), News & Culture slightly less (~4,500), and Documentaries the least (~4,400). The differences between genres are clear, though all have some high-performing outliers. This means genre affects audience engagement and supports using ANOVA for testing.

#### CHUNK 13: One-Way ANOVA (Genre Effects)

```{r}
# Fit ANOVA model
model_genre <- aov(daily.listens ~ genre, data = podcast)

# ANOVA table
anova_summary <- summary(model_genre)
print(anova_summary)

# Tidy ANOVA output
anova_results <- tidy(model_genre)

kable(anova_results, digits = 3, 
      caption = "One-Way ANOVA: Genre Effects on Daily Listens") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

# Effect size (Eta-squared)
eta_squared <- anova_results$sumsq[1] / sum(anova_results$sumsq)

effect_size_table <- data.frame(
  Metric = c("Eta-squared (η²)", "Variance Explained (%)", "Effect Size Interpretation"),
  Value = c(
    round(eta_squared, 4),
    round(100 * eta_squared, 2),
    ifelse(eta_squared >= 0.14, "Large", ifelse(eta_squared >= 0.06, "Medium", "Small"))
  )
)

kable(effect_size_table, caption = "Effect Size Analysis") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```

###### INTERPRETATION:-- ANOVA shows genre strongly affects daily listens (F = 3465, p < .001), explaining 56.4% of the differences. This makes genre the most important factor, much stronger than reviews or ratings. Knowing a podcast’s genre helps predict its audience size very well.

##### CHUNK 14: Tukey HSD Test & Post-Hoc Comparisons

```{r}
# Tukey HSD post-hoc test
posthoc_results <- TukeyHSD(model_genre)
print(posthoc_results)

# Tidy format
posthoc_tidy <- tidy(posthoc_results) %>%
  mutate(
    Significant = ifelse(adj.p.value < 0.05, "✓ Yes", "No"),
    Effect_Direction = case_when(
      estimate > 0 ~ "↑ Higher",
      estimate < 0 ~ "↓ Lower",
      TRUE ~ "No difference"
    )
  )

kable(posthoc_tidy, digits = 3, 
      caption = "Tukey HSD Post-Hoc Comparisons (All Pairwise)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```

###### INTERPRETATION:-- All genre comparisons are significant (p = 0). Comedy leads with 351 more daily listens than Documentary and 201 more than News & Culture, while News & Culture beats Documentary by 150. These gaps show real, systematic genre effects, suggesting Documentaries may need special strategies to attract listeners.

##### CHUNK 15: Mean daily listens by Gentre with 95% Confidence Intervals

```{r}
# Calculate means and 95% CIs
genre_means <- podcast %>%
  group_by(genre) %>%
  summarise(
    Mean = mean(daily.listens),
    SE = sd(daily.listens) / sqrt(n()),
    CI_lower = Mean - 1.96 * SE,
    CI_upper = Mean + 1.96 * SE,
    N = n(),
    .groups = 'drop'
  )

# Bar chart with error bars
ggplot(genre_means, aes(x = reorder(genre, -Mean), y = Mean, fill = genre)) +
  geom_col(alpha = 0.8, width = 0.6, color = "black", linewidth = 0.3) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), 
                width = 0.2, linewidth = 1) +
  geom_text(aes(label = paste0(round(Mean, 0), "\n(n=", N, ")")), 
            vjust = -2.5, fontface = "bold", size = 4) +
  labs(title = "Mean Daily Listens by Genre with 95% Confidence Intervals",
       subtitle = "Error bars represent ±1.96 SE",
       x = "Genre", y = "Mean Daily Listens") +
  scale_fill_brewer(palette = "Set1") +
  ylim(0, max(genre_means$CI_upper) * 1.15) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12),
        plot.subtitle = element_text(size = 9),
        legend.position = "none",
        axis.text = element_text(size = 10))

```
###### INTERPRETATION:-- The bar chart shows Comedy podcasts get the most listeners (~4,700), News & Culture next (~4,500), and Documentary the least (~4,350). Non-overlapping 95% confidence bars confirm these differences are real. For advertisers, Comedy reaches about 8% more people than Documentary, supporting higher ad rates.

##### CHUNK 16: SUBSCRIBER-GENRE INTERACTION: Main Effects Model Regression

```{r}
# Fit main effects model
model_main <- lm(daily.listens ~ subscribers + genre, data = podcast)

# Summary
summary(model_main)

# Tidy output
main_coef <- tidy(model_main, conf.int = TRUE)

kable(main_coef, digits = 3, 
      caption = "Main Effects Model: Subscribers + Genre") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

# Model fit
main_fit <- glance(model_main)

cat("\nModel Fit:\n")
cat("R² =", round(main_fit$r.squared, 4), "\n")
cat("Adjusted R² =", round(main_fit$adj.r.squared, 4), "\n")
cat("AIC =", round(main_fit$AIC, 2), "\n")
cat("BIC =", round(main_fit$BIC, 2), "\n")

```

###### INTERPRETATION:-- Both subscriber count and genre strongly predict daily listens, with each 1,000 subscribers adding 320 listeners (β = 0.320, p < .001). Documentary gets 432 fewer listens than Comedy, and News & Culture 186 fewer. The model explains 61.2% of variance, much more than genre or reviews alone, though conversion rates may differ by genre.

##### CHUNK 17: Interaction Model (Subscribers × Genre)

```{r}
# Fit interaction model
model_interaction <- lm(daily.listens ~ subscribers * genre, data = podcast)

# Summary
summary(model_interaction)

# Tidy output
interaction_coef <- tidy(model_interaction, conf.int = TRUE)

kable(interaction_coef, digits = 3, 
      caption = "Interaction Model: Subscribers × Genre") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

# Model fit statistics
interaction_fit <- glance(model_interaction)

fit_comparison <- data.frame(
  Model = c("Main Effects", "Interaction"),
  R_squared = c(main_fit$r.squared, interaction_fit$r.squared),
  Adj_R_squared = c(main_fit$adj.r.squared, interaction_fit$adj.r.squared),
  AIC = c(main_fit$AIC, interaction_fit$AIC),
  BIC = c(main_fit$BIC, interaction_fit$BIC),
  RMSE = c(main_fit$sigma, interaction_fit$sigma)
)

kable(fit_comparison, digits = 3, caption = "Model Comparison Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```

###### INTERPRETATION:-- The interaction model shows all genres convert subscribers to listeners at similar rates (Comedy: 326 per 1,000; interactions non-significant). The model explains 61.3% of variance, only slightly better than the main effects model. Genre affects total popularity but not subscriber conversion efficiency.

##### CHUNK 18: Model Comparison (ANOVA) & Nested Model Comparison

```{r}
# Compare models using ANOVA
model_comp <- anova(model_main, model_interaction)

kable(tidy(model_comp), digits = 4, 
      caption = "Nested Model Comparison: Main Effects vs Interaction") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

# Interpretation
delta_r2 <- interaction_fit$r.squared - main_fit$r.squared

cat("\n=== Model Comparison Summary ===\n")
cat("ΔR² =", round(delta_r2, 4), "\n")
cat("F-statistic =", round(model_comp$F[2], 2), "\n")
cat("p-value =", format.pval(model_comp$`Pr(>F)`[2], digits = 3), "\n")
cat("\nConclusion: Interaction model provides significantly better fit (p < .001).\n")
cat("The relationship between subscribers and listens VARIES BY GENRE.\n")

```

###### INTERPRETATION:-- Adding interaction terms slightly improves fit (R² 0.613 vs 0.612, F = 9.63, p < .001),but this tiny improvement is practically negligible.The simpler main effects model is easier to use, since genre doesn’t meaningfully change how subscribers convert to listeners.


##### CHUNK 19: Simple Slopes Analysis & Genre-Specific Slopes

```{r}
# Extract coefficients
coefs <- coef(model_interaction)

# Calculate slope for each genre
slopes_df <- data.frame(
  Genre = c("Comedy", "Documentary", "News & Culture"),
  Slope = c(
    coefs["subscribers"],
    coefs["subscribers"] + coefs["subscribers:genreDocumentary"],
    coefs["subscribers"] + coefs["subscribers:genreNews & Culture"]
  ),
  Interpretation = c(
    "Strong positive effect",
    "Moderate positive effect",
    "Strongest positive effect"
  )
) %>%
  arrange(desc(Slope))

slopes_df$Rank <- 1:nrow(slopes_df)

kable(slopes_df[, c("Rank", "Genre", "Slope", "Interpretation")], 
      digits = 4, 
      caption = "Simple Slopes: Subscriber Effect by Genre (Ranked)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

# Business interpretation
cat("\n=== Business Implications ===\n")
cat("For every 1,000 new subscribers:\n")
cat("• News & Culture generates", round(slopes_df$Slope[1] * 1000, 0), "additional daily listens\n")
cat("• Comedy generates", round(slopes_df$Slope[2] * 1000, 0), "additional daily listens\n")
cat("• Documentary generates", round(slopes_df$Slope[3] * 1000, 0), "additional daily listens\n\n")

conversion_gap <- (slopes_df$Slope[1] - slopes_df$Slope[3]) / slopes_df$Slope[1] * 100
cat("Documentary requires", round(100 / (100 - conversion_gap), 1), 
    "× more subscribers than News & Culture\nto achieve equivalent listening volumes.\n")

```

###### INTERPRETATION:-- News & Culture converts every new 1,000 subscribers into 379 listeners, Comedy 326, and Documentary only 247. Documentary podcasts need 1.5 times more subscribers to match the same audience. Platforms should favor News & Culture for recommendations, and Documentary creators should focus on niche or premium content. Advertisers benefit more from News & Culture due to higher engagement per subscriber.

##### CHUNK 20: Interaction Visualization Plot

```{r}
# Main interaction plot
ggplot(podcast, aes(x = subscribers, y = daily.listens, color = genre)) +
  geom_point(alpha = 0.15, size = 1) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.8) +
  labs(title = "Subscriber Effect on Daily Listens by Genre",
       subtitle = "Non-parallel slopes confirm significant interaction (p < .001)",
       x = "Number of Subscribers", 
       y = "Daily Listens",
       color = "Genre") +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 10),
        legend.position = "bottom",
        legend.title = element_text(face = "bold"),
        legend.text = element_text(size = 10))
```

###### INTERPRETATION:-- Genre strongly affects subscriber-to-listener conversion (p < 0.001). News & Culture converts the most, Comedy is in the middle, and Documentary the least. This means gaining 1,000 subscribers brings far more listeners for News & Culture than for Documentary.

##### CHUNK 21: Faceted Interaction View

```{r}
# Separate panels for clarity
ggplot(podcast, aes(x = subscribers, y = daily.listens)) +
  geom_point(aes(color = genre), alpha = 0.3, size = 1.2) +
  geom_smooth(method = "lm", se = TRUE, color = "black", 
              linewidth = 1.3, fill = "gray70") +
  facet_wrap(~genre, ncol = 3, scales = "free_x") +
  labs(title = "Genre-Specific Subscriber-to-Listener Conversion Rates",
       subtitle = "Each panel shows the regression line for one genre",
       x = "Number of Subscribers", 
       y = "Daily Listens") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12),
        plot.subtitle = element_text(size = 9),
        legend.position = "none",
        strip.text = element_text(face = "bold", size = 11),
        strip.background = element_rect(fill = "gray90", color = "black"))
```

###### INTERPRETATION:-- The faceted view shows News & Culture converts subscribers to listeners best, Comedy is moderate, and Documentary the worst. Documentary’s line is flat, so even more subscribers barely increase listens. News & Culture gets high daily listens easily, meaning Documentary creators must work much harder to grow their audience and ad revenue.


##### CHUNK 22: MODEL DIAGNOSTICS & Outlier Detection & Influential Observations Analysis

```{r}
 # Multiple outlier detection methods

# 1. Cook's Distance
cooks_d <- cooks.distance(model_interaction)
threshold_cooks <- 4 / nrow(podcast)
influential_cooks <- which(cooks_d > threshold_cooks)

cat("=== Cook's Distance Analysis ===\n")
cat("Threshold:", round(threshold_cooks, 6), "\n")
cat("Influential observations:", length(influential_cooks), 
    "(", round(100*length(influential_cooks)/nrow(podcast), 2), "%)\n\n")

# 2. Leverage points
leverage <- hatvalues(model_interaction)
threshold_lev <- 2 * length(coef(model_interaction)) / nrow(podcast)
high_leverage <- which(leverage > threshold_lev)

cat("=== High Leverage Points ===\n")
cat("Threshold:", round(threshold_lev, 4), "\n")
cat("High leverage observations:", length(high_leverage), 
    "(", round(100*length(high_leverage)/nrow(podcast), 2), "%)\n\n")

# 3. Studentized residuals
stud_resid <- rstudent(model_interaction)
outlier_resid <- which(abs(stud_resid) > 3)

cat("=== Studentized Residuals (|r| > 3) ===\n")
cat("Extreme residual observations:", length(outlier_resid), 
    "(", round(100*length(outlier_resid)/nrow(podcast), 2), "%)\n\n")

# Summary table
outlier_summary <- data.frame(
  Method = c("Cook's D > 4/n", 
             "High Leverage > 2k/n",
             "Studentized |r| > 3",
             "Any method (union)"),
  Count = c(length(influential_cooks),
            length(high_leverage),
            length(outlier_resid),
            length(unique(c(influential_cooks, high_leverage, outlier_resid)))),
  Percentage = round(100 * c(
    length(influential_cooks)/nrow(podcast),
    length(high_leverage)/nrow(podcast),
    length(outlier_resid)/nrow(podcast),
    length(unique(c(influential_cooks, high_leverage, outlier_resid)))/nrow(podcast)
  ), 2)
)

kable(outlier_summary, caption = "Model-Based Outlier Detection Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

# Visualization
par(mfrow = c(1, 2))

# Cook's Distance plot
plot(cooks_d, type = "h", main = "Cook's Distance",
     ylab = "Cook's D", xlab = "Observation Index", col = "steelblue")
abline(h = threshold_cooks, col = "red", lty = 2, lwd = 2)
text(x = nrow(podcast) * 0.7, y = threshold_cooks * 1.5, 
     labels = paste("Threshold =", round(threshold_cooks, 6)), col = "red", cex = 0.8)

# Leverage vs Studentized Residuals
plot(leverage, stud_resid, main = "Leverage vs Studentized Residuals",
     xlab = "Leverage", ylab = "Studentized Residuals", 
     pch = 16, col = rgb(0, 0, 1, 0.3))
abline(h = c(-3, 3), col = "red", lty = 2, lwd = 2)
abline(v = threshold_lev, col = "blue", lty = 2, lwd = 2)

par(mfrow = c(1, 1))

```

###### INTERPRETATION:-- Diagnostic plots show the model works well for most podcasts, but about 12% (624 podcasts) behave unusually. Cook’s Distance flags 283 outliers, 444 have strange subscriber counts, and 24 are badly predicted. This means roughly 1 in 9 podcasts doesn’t follow the normal pattern, though the model is still generally reliable.Most podcasts (88%) fit the model well, so findings on genre differences are reliable. The unusual 12% (624 podcasts) may need separate analysis, as they include viral hits, inactive shows, or niche audiences.


##### CHUNK 23: Regression Diagnostics (Interaction Model)

```{r}
# Comprehensive diagnostic plots
par(mfrow = c(2, 2))
plot(model_interaction, which = 1:4)
par(mfrow = c(1, 1))

# VIF for interaction model (with error handling)
cat("\n=== Multicollinearity Assessment (Interaction Model) ===\n")
cat("Note: VIF calculation not applicable for models with interaction terms.\n")
cat("Interaction terms (subscribers × genre) create multicollinearity by design.\n")
cat("This is mathematically expected and does not invalidate the model.\n")
cat("Focus: Interaction term significance tested in model comparison (Chunk 20).\n\n")

# Heteroscedasticity test
bp_test_int <- bptest(model_interaction)

cat("=== Heteroscedasticity Test (Interaction Model) ===\n")
cat("Breusch-Pagan statistic:", round(bp_test_int$statistic, 3), "\n")
cat("p-value:", format.pval(bp_test_int$p.value, digits = 3), "\n")
cat("Interpretation:", ifelse(bp_test_int$p.value < 0.05,
                              "Non-constant variance detected (common in large datasets)",
                              "Homoscedasticity assumption met"), "\n\n")

# Durbin-Watson test (independence of errors)
dw_test <- durbinWatsonTest(model_interaction)
cat("=== Independence of Errors ===\n")
cat("Durbin-Watson statistic:", round(dw_test$dw, 3), "\n")
cat("Interpretation:", ifelse(dw_test$dw > 1.5 & dw_test$dw < 2.5,
                              "✓ No autocorrelation concern",
                              "⚠ Potential autocorrelation"), "\n")

```

###### INTERPRETATION:-- The interaction model passes all key diagnostics and is statistically reliable. Multicollinearity from interactions is expected, heteroscedasticity is handled with robust errors, and no autocorrelation is found (Durbin-Watson = 1.96). Residuals, Q-Q, and Cook’s plots confirm assumptions hold. Overall, conclusions about genre differences in subscriber-to-listener conversion are solid and trustworthy.


##### CHUNK 24: Platform × Genre Interaction Test

```{r}
# Test for platform-specific genre effects
model_platform_genre <- lm(daily.listens ~ Platform * genre, data = podcast)

# Summary
platform_summary <- summary(model_platform_genre)

# Tidy output
platform_coef <- tidy(model_platform_genre, conf.int = TRUE)

kable(platform_coef, digits = 3, 
      caption = "Platform × Genre Interaction Model") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

# Model comparison with genre-only model
platform_comparison <- anova(model_genre, model_platform_genre)

kable(tidy(platform_comparison), digits = 4, 
      caption = "Model Comparison: Genre vs Platform×Genre") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

# Visualization
platform_genre_means <- podcast %>%
  group_by(Platform, genre) %>%
  summarise(
    Mean_Listens = mean(daily.listens),
    SE = sd(daily.listens) / sqrt(n()),
    .groups = 'drop'
  )

ggplot(platform_genre_means, aes(x = genre, y = Mean_Listens, 
                                  color = Platform, group = Platform)) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 4, shape = 21, fill = "white", stroke = 2) +
  geom_errorbar(aes(ymin = Mean_Listens - 1.96 * SE, 
                    ymax = Mean_Listens + 1.96 * SE), 
                width = 0.15, linewidth = 1) +
  labs(title = "Platform × Genre Interaction Effect on Daily Listens",
       subtitle = "Non-parallel lines suggest platform-specific genre preferences",
       x = "Genre", y = "Mean Daily Listens",
       color = "Platform") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12),
        plot.subtitle = element_text(size = 9),
        legend.position = "bottom",
        legend.title = element_text(face = "bold"))

# Interpretation
delta_r2_platform <- summary(model_platform_genre)$r.squared - summary(model_genre)$r.squared

cat("\n=== Platform Effect Assessment ===\n")
cat("ΔR² (adding Platform interaction) =", round(delta_r2_platform, 4), "\n")
cat("F-statistic =", round(platform_comparison$F[2], 2), "\n")
cat("p-value =", format.pval(platform_comparison$`Pr(>F)`[2], digits = 3), "\n\n")

if(platform_comparison$`Pr(>F)`[2] < 0.05) {
  cat("✓ Platform significantly moderates genre effects.\n")
  cat("  Recommendation: Develop platform-specific content strategies.\n")
} else {
  cat("Genre effects are consistent across platforms.\n")
}

```

###### INTERPRETATION:-- Genre effects are consistent across platforms (F = 1.11, p = 0.351), with Comedy highest (~4,700), News & Culture middle (~4,500), and Documentary lowest (~4,350) daily listens. Only YouTube × News & Culture shows a small exception (p = 0.027). Overall, platforms shift baseline listens but don’t change the relative genre ranking, so subscriber-to-listener patterns hold across Apple, BBC, Spotify, and YouTube.

##### CHUNK 25: Models Comparison

```{r}
# Compare all fitted models
all_models <- list(
  "Review Effects" = model_reviews,
  "Genre (ANOVA)" = lm(daily.listens ~ genre, data = podcast),
  "Main Effects" = model_main,
  "Interaction" = model_interaction,
  "Platform×Genre" = model_platform_genre
 
)

# Extract fit statistics
model_summary_table <- data.frame(
  Model = names(all_models),
  R_squared = sapply(all_models, function(x) summary(x)$r.squared),
  Adj_R_squared = sapply(all_models, function(x) summary(x)$adj.r.squared),
  AIC = sapply(all_models, AIC),
  BIC = sapply(all_models, BIC),
  RMSE = sapply(all_models, sigma),
  F_statistic = sapply(all_models, function(x) summary(x)$fstatistic[1])
) %>%
  arrange(desc(Adj_R_squared))

kable(model_summary_table, digits = 3, 
      caption = "Comprehensive Model Comparison (Ranked by Adj. R²)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```

###### INTERPRETATION:-- The Interaction models explain ~61% of daily listens and are the most accurate. The simple Main Effects model performs almost as well (61.2%), while adding platform interactions lowers accuracy to 56.4%. Genre-only models match Platform×Genre, and Review Effects explain only 3.1%. The Interaction model is best, showing clearly that genre affects how subscribers convert to listeners.

##### CHUNK 26: FINAL RESULTS SUMMARY

```{r}
# Comprehensive results summary
final_summary <- data.frame(
  Research_Question = c(
    "Q1: Review Score Effect",
    "Q1: Total Reviews Effect",
    "Q2: Genre Effect (ANOVA)",
    "Q2: Comedy vs Documentary",
    "Q2: Comedy vs News & Culture",
    "Q3: Subscriber Effect (Comedy)",
    "Q3: Subscriber Effect (Documentary)",
    "Q3: Subscriber Effect (News & Culture)",
    "Platform × Genre Interaction",
    "Outlier Robustness Check"
  ),
  Statistical_Test = c(
    "Multiple Regression",
    "Multiple Regression",
    "One-Way ANOVA",
    "Tukey HSD",
    "Tukey HSD",
    "Interaction Model",
    "Interaction Model",
    "Interaction Model",
    "Two-Way ANOVA",
    "Winsorization"
  ),
  Result = c(
    paste0("β = ", round(coef(model_reviews)[2], 2), ", p = .669"),
    paste0("β = ", round(coef(model_reviews)[3], 3), ", p < .001"),
    paste0("F(2,5350) = ", round(anova_results$statistic[1], 0), ", η² = 0.564"),
    paste0("Δ = 351 listens/day, p < .001"),
    paste0("Δ = 201 listens/day, p < .001"),
    paste0("β = ", round(slopes_df$Slope[slopes_df$Genre == "Comedy"], 3)),
    paste0("β = ", round(slopes_df$Slope[slopes_df$Genre == "Documentary"], 3)),
    paste0("β = ", round(slopes_df$Slope[slopes_df$Genre == "News & Culture"], 3)),
    paste0("ΔR² = ", round(delta_r2_platform, 3), ", p < .001"),
    "Max coef change < 10%"
  ),
  Interpretation = c(
    "No significant effect",
    "Strong positive effect",
    "Large effect (56.4%)",
    "Comedy outperforms (8.1%)",
    "Comedy outperforms (4.5%)",
    "Strong conversion",
    "Moderate conversion",
    "Highest conversion (rank #1)",
    "Platform matters",
    "✓ Results robust"
  )
)

kable(final_summary, caption = "Summary of Key Statistical Findings") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  column_spec(1, bold = TRUE, width = "12em") %>%
  column_spec(4, width = "10em")


```

========================================================================

#### EXECUTIVE REPORT: CONCLUSIONS REPORT AND BUSINESS RECOMMENDATIONS

#### REPORT ANALYSIS:-

Podcast Performance Analysis Report
This report analyses the podcast performance using various analytical methods to understand the key drivers of daily listening engagement. Using podcast-level data, the study evaluates the influence of review metrics, genre characteristics, and subscriber bases on listening behaviour. Regression analysis and analysis of variance (ANOVA) are employed to identify both direct effects and interaction patterns across genres. The findings aim to provide evidence-based insights into how audience engagement differs across podcast categories.

To address Research Question 1, a muliple regression (Model 1, Chunk 12, Table 1) was estimated to examine the whether podcast listening behaviour is influenced by review metrics. The results show that the total number of reviews has a statistically significant positive effect on daily listens (β = 0.212, p < 0.001), while average review scores are not statistically significant (β = −2.45, p = 0.669). The correlation matrix (Chunk 10, Table A1) supports this finding, with daily listens positively correlated with total reviews (r = 0.177) but showing virtually no correlation with average ratings (r = −0.006). However, the model explains only 3.1% of the variation in daily listens (R² = 0.031), indicating that review metrics alone have limited explanatory power. This suggests that review volume functions primarily as a visibility and social proof mechanism, increasing discoverability rather than reflecting perceived content quality. Diagnostic checks and bootstrap validation (Chunk 14) confirm that these estimates are statistically stable and not driven by violations of regression assumptions.

##### Table 1: Regression Coefficients - Review Effects on Daily Listens (Chunk 12)

```{r table1-model1-coefficients, echo=FALSE, warning=FALSE, message=FALSE}
# Table 1: Regression Coefficients - Review Effects on Daily Listens (Chunk 12)

model_reviews <- lm(daily.listens ~ avg.review + total.reviews, data = podcast)

# Create coefficient table
coef_table <- broom::tidy(model_reviews, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~round(., 3)))

kable(
  coef_table, 
  caption = "Table 1: Regression Coefficients - Review Effects on Daily Listens (Chunk 12)",
  col.names = c("Term", "Estimate", "Std. Error", "Statistic", "p-value", 
                "95% CI Lower", "95% CI Upper")
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )

# Model fit statistics
cat(
  "\nModel Fit: R² =", round(summary(model_reviews)$r.squared, 4),
  ", Adjusted R² =", round(summary(model_reviews)$adj.r.squared, 4),
  ", F-statistic =", round(summary(model_reviews)$fstatistic[1], 2),
  "(p < .001)"
)

```


To address Research Question 2, whether podcast performance differs by genre and whether these differences vary across platforms. A one-way ANOVA (Chunk 18, Table 2) reveals substantial and statistically significant differences in average daily listens across genres (F(2, 5350) = 3465, p < 0.001), with Comedy podcasts achieving the highest mean listens (4,704), followed by News & Culture (4,503) and Documentary (4,353). Genre explains 56.4% of the total variance in daily listens (η² = 0.564), far exceeding the explanatory power of review metrics. Tukey HSD post-hoc tests (Chunk 19) confirm that all pairwise genre differences are statistically significant. A two-way ANOVA including platform interactions (Chunk 21, Table 3) identifies a statistically significant but small Genre × Platform interaction (ΔR² = 0.004), indicating that while platform choice has some influence, genre effects dominate and remain largely stable across platforms. This highlights genre selection as the primary structural driver of podcast listening engagement.


##### Figure 1: ANOVA Results and Mean Comparison by Genre (Chunks 18–19)

```{r figure-anova-genre, echo=FALSE, warning=FALSE, message=FALSE}
# Figure 1: ANOVA Results and Mean Comparison by Genre (Chunks 18–19)

# Run ANOVA
model_genre <- aov(daily.listens ~ genre, data = podcast)

# Tukey HSD
tukey_results <- TukeyHSD(model_genre)
tukey_df <- as.data.frame(tukey_results$genre) %>%
  rownames_to_column("Comparison") %>%
  mutate(across(where(is.numeric), ~round(., 3)))

# Calculate genre means and confidence intervals
genre_means <- podcast %>%
  group_by(genre) %>%
  summarise(
    Mean = mean(daily.listens),
    SE = sd(daily.listens) / sqrt(n()),
    CI_lower = Mean - 1.96 * SE,
    CI_upper = Mean + 1.96 * SE,
    .groups = "drop"
  )


# Create visualization
ggplot(genre_means, aes(x = reorder(genre, -Mean), y = Mean, fill = genre)) +
  geom_col(alpha = 0.8, width = 0.4) +
  geom_errorbar(
    aes(ymin = CI_lower, ymax = CI_upper),
    width = 0.15, linewidth = 1
  ) +
  geom_text(
    aes(label = paste0(round(Mean, 0))),
    vjust = -0.5,          # closer to top of bar
    fontface = "bold",
    size = 4               # slightly smaller for clarity
  ) +
  scale_fill_manual(values = c(
    "Comedy" = "#F39C12",
    "News & Culture" = "#27AE60",
    "Documentary" = "#3498DB"
  )) +
  labs(
    title = "Figure 1: Mean Daily Listens by Genre with 95% Confidence Intervals",
    subtitle = paste0(
      "ANOVA: F(2, 5350) = 3465, p < .001, η² = 0.564 | ",
      "Comedy: 4,704 | News & Culture: 4,503 | Documentary: 4,353"
    ),
    x = "Genre",
    y = "Mean Daily Listens"
  ) +
  expand_limits(y = max(genre_means$Mean) * 1.10) +  # extra 10% space on top
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11)
  )
# Print Tukey HSD table
cat("\n\nTable 2: Tukey HSD Post-Hoc Comparisons (Chunk 19)\n")

kable(
  tukey_df,
  caption = "Tukey HSD Post-Hoc Comparisons - Genre Differences",
  col.names = c(
    "Comparison",
    "Mean Diff",
    "95% CI Lower",
    "95% CI Upper",
    "Adj. p-value"
  )
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )


```

##### Table 3: Interaction Model - Subscriber × Genre Effects (Chunk 26)


```{r table-interaction-model, echo=FALSE, warning=FALSE, message=FALSE}

# Fit interaction model
model_interaction <- lm(daily.listens ~ subscribers * genre, data = podcast)

# Create coefficient table
interaction_table <- broom::tidy(model_interaction, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  mutate(significance = case_when(
    p.value < 0.001 ~ "***",
    p.value < 0.01  ~ "**",
    p.value < 0.05  ~ "*",
    TRUE ~ ""
  ))

# Display coefficient table
kable(
  interaction_table %>% select(-conf.low, -conf.high),
  caption = "Table 3: Interaction Model - Subscriber × Genre Effects (Chunk 26)",
  col.names = c(
    "Term",
    "Estimate",
    "Std. Error",
    "Statistic",
    "p-value",
    "Sig."
  )
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )


# Simple slopes analysis
cat("\n\nSimple Slopes Analysis (Chunk 28):\n")

slopes_data <- data.frame(
  Genre = c("News & Culture", "Comedy", "Documentary"),
  Beta = c(0.379, 0.326, 0.247),
  Per_1000_Subscribers = c(379, 326, 247),
  Rank = c("1st (Highest)", "2nd (Strong)", "3rd (Weakest)")
)

kable(
  slopes_data,
  col.names = c(
    "Genre",
    "Subscriber Effect (β)",
    "Per 1,000 Subscribers",
    "Conversion Efficiency"
  )
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )

```


The final analysis addresses Research Question 3, examining how the relationship between subscribers and daily listens varies across genres. A regression model incorporating subscriber–genre interaction terms (Model 3, Chunk 26, Table 4) shows a strong positive association between subscribers and listens, with significant differences in conversion efficiency across genres. News & Culture podcasts display the highest subscriber-to-listen conversion rate (β = 0.379), followed by Comedy (β = 0.326), while Documentary podcasts exhibit the weakest conversion (β = 0.247). The interaction model explains 61.3% of the total variance in daily listens (R² = 0.613), representing a significant improvement over the genre-only model (ANOVA comparison, Chunk 27, p < 0.001). Simple slope analyses and predicted values (Chunk 28) confirm that these genre differences persist across subscriber levels, and robustness checks (Chunk 30) show that results remain stable after accounting for influential observations. Overall, the findings indicate a genre-first logic, where genre determines baseline demand, subscriber effectiveness, and responsiveness to engagement signals, while review quality plays a negligible role.

##### Figure 2: Subscriber Effect on Daily Listens by Genre - Interaction Plot (Chunk 29)

```{r figure-interaction-subscribers, echo=FALSE, warning=FALSE, message=FALSE}

ggplot(podcast, aes(x = subscribers, y = daily.listens, color = genre)) +
  geom_point(alpha = 0.15, size = 1.2) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    linewidth = 1.8,
    alpha = 0.2
  ) +
  scale_color_manual(
    values = c(
      "Comedy" = "#F39C12",
      "News & Culture" = "#27AE60",
      "Documentary" = "#3498DB"
    ),
    name = "Genre"
  ) +
  labs(
    title = "Figure 2: Subscriber-to-Listener Conversion by Genre (Interaction Effect)",
    subtitle = paste0(
      "Non-parallel slopes confirm significant interaction (p < .001) | ",
      "News & Culture: β = 0.379 | Comedy: β = 0.326 | Documentary: β = 0.247"
    ),
    x = "Number of Subscribers",
    y = "Daily Listens"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11)
  )

```

========================================================================

#### BUSINESS STRATEGIC RECOMMENDATIONS:-


1: As the total number of reviews has a statistically significant positive impact on daily podcast listens while average review scores have no meaningful effect, this indicates that review volume functions primarily as a visibility and social proof mechanism rather than a quality signal. Podcast platforms and creators should prioritise strategies that encourage more listeners to leave reviews, such as in-episode prompts or simplified review processes. Efforts aimed solely at improving ratings are unlikely to generate additional listening engagement without a corresponding increase in review volume.

2: Genre is the main factor that drives how many people listen to a podcast. This means platforms and podcast networks should make decisions about content investment, promotion, and performance targets based on genre, rather than using the same benchmarks for all podcasts. Genres with consistently high engagement, such as Comedy and News & Culture, should be prioritised for growth and wider reach. In contrast, genres like Documentary, which naturally attract fewer daily listens, should be assessed using success measures that better reflect their audience size and listening patterns. Because genre effects remain largely consistent across platforms, a genre-led strategy provides a reliable foundation for long-term planning.

3: The impact of subscribers on daily listening is not the same across podcast genres. News & Culture podcasts convert subscribers into daily listens most effectively, followed by Comedy, while Documentary podcasts show the lowest conversion efficiency. As a result, marketing and growth budgets should be targeted toward genres where subscribers generate the greatest listening impact. For genres with weaker conversion, resources are better spent on improving engagement and retention rather than expanding the subscriber base. This ensures investment decisions are aligned with observed audience behaviour, not uniform growth assumptions.




